---
title: "Analysis of longitudinal data"
---

```{r}
knitr::opts_chunk$set(message = FALSE)
date()
```

```{r}
# BRING IN THE DATA

# access packages readr and dplyr
library(readr); library(dplyr)

# set the working directory
setwd("C:/Users/yona2/Desktop/FT/IODS2022/RStudio/IODS-project/data")

# load the datasets
RATSL <- read.csv("RATSL.csv")
BPRSL <- read.csv("BPRSL.csv")

# check structure
str(RATSL)
str(BPRSL)

# factor variables ID and Group in RATSL
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

# factor variables treatment and subject in BPRSL
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

# check structure again
str(RATSL); summary(RATSL)
str(BPRSL); summary(BPRSL)

```

The RATS data contain information on 16 rats that were assigned to 3 treatment groups. Weight was measured 11 times, at baseline on day 1 (WD1) and about once a week thereafter up to day 64 (WD64). During the study, the weight varied from 225 to 628 g.

The BPRS data contain information on 40 subjects that were assigned to 2 treatment groups, 20 subjects in each. Response to treatment was measured 9 times by the Brief psychiatric rating scale (BPRS), at baseline (week0) and weekly thereafter up to 8 weeks (week1-week8). During the study, BPRS varied from 18 to 95.


```{r}
# ANALYSING RATS DATA

# access the package ggplot2
library(ggplot2)

# draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, times=2)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

# standardise the variable bprs
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# check structure and summary anew
str(RATSL); summary(RATSL)

# plot again with the standardized bprs
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, times=2)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized bprs")

```

Standardization decreases considerably the variability of weight as expected. In the plots, tracking can be detected and the phenomenon is somewhat clearer in the standardized plots. It is also notable that the variance of weight seems to decrease somewhat along the follow-up in all treatment groups even though it is less evident in group 1. There seems to be weight gain during time in all groups and especially in groups 2 and 3. In group 2, one rat seems to be in a league of its own.


```{r}
# number of subjects (per group):
n <- 8   # note: group 1 has 8 rats, and groups 2 and 3 4 rats

library(tidyr)
# summary data with mean and standard error of weight by group and time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()

# correction to se in groups 2 and 3
group <- RATSS$Group %>% as.numeric
RATSS$se <- ifelse(group > 1, sqrt(2)*RATSS$se, RATSS$se)

# plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

In the plot, mean (SE) values of weight are shown at each timepoint by treatment group. The variances are larger in groups 2 and 3 compared with group 1. Variance may decrease a little in group 3 over time but in other groups there is not much change. The seems to be weight gain in all groups but the slopes look bigger in groups 2 and 3.


```{r}
# create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0)
RATSL10S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight)) %>%
  ungroup()

# check structure and summary of the new object
str(RATSL10S); summary(RATSL10S)

# draw a boxplot of the mean versus treatment
ggplot(RATSL10S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 1-64")

# create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL10S1 <- RATSL10S %>% filter(mean < 580)

# check structure and summary of the object
str(RATSL10S1); summary(RATSL10S1)

# draw a boxplot of the mean versus treatment again
ggplot(RATSL10S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 1-64")

```

When comparing the overall means it can be seen that there is an outlier in group 2 (mean weight > 590). Filtering the outlier changes the boxplot of group 2. There is a remarkable difference between groups but it is mostly explained by the difference at baseline.


```{r}
# t-test can be used only for 2 groups, so fit a basic linear model instead
fit1 <- lm(mean ~ Group, data = RATSL10S1)
summary(fit1)

# compute the analysis of variance table for the fitted model with anova()
anova(fit1)

# read the RATS data in order to get baseline values
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

# add the baseline from the original data as a new variable to the summary data
RATSL10S2 <- RATSL10S %>%
  mutate(baseline = RATS$WD1) %>%  # note: the number of observations needs to be the same, hence filtering anew
  filter(mean < 590)

# fit the linear model with the mean as the response 
fit2 <- lm(mean ~ baseline + Group, data = RATSL10S2)
summary(fit2)

# compute the analysis of variance table for the fitted model with anova()
anova(fit2)

```

As expected, in the base model, the mean overall weight differs significantly between groups and the results are in concordance with the previous plot. The model explains 99% of the variation in mean overall weight. The mean overall weight in group 2 is 187 g higher and in group 262 g higher compared with group 1. But the result does not tell anything about the differences in weight gain between the groups during the experiment.

Baseline weight is a statistically significant addition to the model: it explains vastly more of the variation than treatment. Yet, there seems to be statistically significant effect by group: the mean overall weight in group 2 is still 89 g higher and in group 113 g higher compared with group 1.

Using the overall mean can be justified if we are interested in the treatment effect along the whole time and not just the end situation. If we are interested only in the final status, especially if we expect that status to be more or less permanent, the more accurate comparison might be to compare the weight status at the end.

Note of caution: in these analyses, the mean overall weight included also the weight at baseline. Because there were 11 measurements, the error it produces to the results is most likely unremarkable.


```{r}
# # ANALYSING BPRS DATA

# Plot the BPRS data (note: each subject needs ty be identified by subject number)
treatment <- BPRSL$treatment %>% as.numeric
BPRSL$subject1 <- ifelse(treatment == 2, 20+as.numeric(BPRSL$subject), as.numeric(BPRSL$subject))
ggplot(BPRSL, aes(x = week, y = bprs, group = subject1, color = subject1)) +
  geom_line(aes(linetype = treatment)) + scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) + 
  scale_y_continuous(name = "BPRS") 

# Check the structure of the data
str(BPRSL)

```

In the plot, there does not seem to be much difference between groups. Tracking seems prevalent and the variation of BPRS seems to decrease by time.


```{r}
# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)

# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject1), data = BPRSL, REML = FALSE)

# print the model summary
summary(BPRS_ref)

```

In the fixed effects model, the observations are modeled as if they are independent from each other. Time (week) is a significant explanatory variable but treatment group is not. The model explains only 19% of variation in BPRS points. The fixed effects estimate for week (-2.3) implies 18 points decrease in 8 weeks which can be considered clinically significant. So time seems to heal but not the treatment.

In the random effects model, the variation between subjects is being taken into account. The fixed effects estimates remain the same but a part of the variance in BPRS points seems to be explained by random effects between subjects. 


```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject1), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```

Taking into account the dependence of BPRS points across time by each subject by adding week in the random effects model, there is still no treatment effect. On the other hand, there seems to be time dependence: when comparing models with and without week AIC decreases and logLik increases, the difference being statistically significant. There are minor changes in the fixed effects estimates.

These models show that large part of the variation between BPRS values can be explained by interindividual and intraindividual random variation. The impression that in this case time heals, not the treatment, becomes stronger. 


```{r}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + week * treatment + (week | subject1), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of RATSL with the observed Weight values
ggplot(BPRSL, aes(x = week, y = bprs, group = subject1)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Observed BPRS points") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)
# Create a new column fitted to RATSL
BPRSL <- BPRSL %>% mutate(Fitted = Fitted)

# draw the plot of RATSL with the Fitted values of weight
ggplot(BPRSL, aes(x = week, y = Fitted, group = subject1)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "top")

```

Adding interaction term for week and treatment into the random effects model does not improve the model, AIC and logLik do not practically change at all. The model is able to fit the observations reasonably well if one is willing to believe that the relationship between BPRS and time is linear; yet, a second-degree polynomial model might be better.

